---
title: 'SDS-192: Mini-Project #02'
author: "Sarah Elghazoly, Abby Doctor"
date: "3/10/2017"
output: html_document
---
##Loading the files
```{r}
library(dplyr)
library(plyr)
load("house_elections.rda")
load("candidates.rda")
load("committees.rda")
load("contributions.rda")
library(ggplot2)
require(treemap)
```

##Number of incumbents who won House positions, number of new runners (for each state)
```{r}
g1 <- house_elections %>% filter(ge_winner == "W") %>% group_by(incumbent)%>% group_by(state)
ggplot(g1, aes(x = state)) +geom_bar(aes(fill = incumbent))
```

##Contribution Amounts for Republican and Democratic parties broken down by year
So far I have total transaction amounts by party
```{r}
##
f2 <- contributions %>% inner_join(candidates, by = "cand_id", na.rm = FALSE) %>% arrange(by = cand_party_affiliation) %>% filter(cand_party_affiliation ==  c("DEM", "REP"))

ggplot(f2, aes(x = cand_state, y = sum(transaction_amt))) + geom_bar(stat = "identity", aes(fill = cand_party_affiliation)) + labs(x = "State", y = "Sum of Contributions") + scale_y_continuous(labels = scales::comma)
```

##Making a Big Data Collection house_elections, candidates
Candidate elections include the names, states, incumbency status, party, number of votes (general), winner status and office
```{r}
##renaming columns in candidates in order to combine candidates and house_elections
candidates <- rename(candidates, c("cand_state" = "state", "cand_name" = "candidate_name", "cand_election_yr" = "year"))

##mutating the chr of house_elections to merge candidates and house_elections by candidate names
house_elections <- house_elections %>% mutate_each(funs(toupper))

##converting chr value house_elections$general_votes to int values
house_elections$general_votes <- as.numeric(house_elections$general_votes)

##inner_join candidate elections
cand_elections <- candidates %>% inner_join(house_elections, by  = c("candidate_name", "state"), na.rm = FALSE)

##cutting cand_elections down to include only necessary columns
cand_elections <- cand_elections %>% select(candidate_name, year, state, incumbent, party, general_votes, ge_winner, cand_office)
```

##Let's look at Arizona House Elections?
```{r}
##Arizona treemap
ariz_cand_elec <- cand_elections %>% filter(state == "AZ", cand_office == "H") %>% mutate(prop_votes = general_votes/sum(general_votes))
treemap(ariz_cand_elec, index = c("party", "candidate_name"), vSize = "general_votes", vColor = c("party", "incumbent"))
```


##Function by State
```{r}
##function that generates a tree graph with a breakdown of candidates and party and number ofvotes in the general election
percentage_breakdown <- function(this_state) {
  g4 <- cand_elections %>% filter(state == this_state, cand_office == "H")
  treemap(g4, index =  c("party", "candidate_name"), vSize = "general_votes", vColor = c("party", "incumbent"))
}

##A few examples
percentage_breakdown(this_state = "IA")
percentage_breakdown(this_state = "GA")
percentage_breakdown(this_state = "FL")
percentage_breakdown(this_state = "CA")
percentage_breakdown(this_state = "AZ")
percentage_breakdown(this_state = "NY")

```


##Choosing an office
```{r}
##breakdown of votes by party for each state
house_values <- cand_elections %>% 
  filter(cand_office == "H") %>%
  arrange(by = party) %>%
  filter
house_values

ggplot(house_values, aes(x = state, y = general_votes)) + geom_bar(stat = "identity", aes(fill = party))+ labs(x = "State", y = "Number of Votes (General)") + scale_y_continuous(labels = scales::comma)
```
